---
title: "Background to R"
author: "David McAllister"
date: "Tuesday, August 05, 2014"
output:
  beamer_presentation: default
  ioslides_presentation: default
---

## Aims and objectives
The aim of this course is to have a better feel for how R works as a language. At the end of this course you should be able to

* Clean and transform data in R
* describe how R's object work
* Understand from the help menu how functions in base R and the packages work
* Use the apply family of functions
* Write simple functions

## Overview

 Time | Topic
------|------ 
0930 to 0945| Introduction
0945 to 1045| Objects
1045 to 1100| Coffee
1100 to 1200| Subsetting
1200 to 1300| Lunch
1300 to 1430| Functions (and factors)
1430 to 1445| Coffee
1445 to 1700| Coding style, gotchas, R Help, other Rs

## Morning in detail
 Time | Topic
------|------ 
0930 to 0945 |Introduction
0945 to 1015 |Objects
1015 to 1045 |working with vectors
1045 to 1100 | **Coffee**
1100 to 1130 |Subsetting vectors
1130 to 1200 |Subsetting lists
1200 to 1300 | **Lunch (not provided)**


## Afternoon in detail

 Time | Topic
------|------ 
1300 to 1330 |Factors and other attributes
1330 to 1400 |Understanding functions
1400 to 1430 |Apply family of functions
1430 to 1445 | **Coffee**
1445 to 1515 |Writing functions
1515 to 1545 |Plyr and reshape2
1545 to 1615 |ggplot2 and other interlopers
1615 to 1700 |Coding style and help menu 

## The S programming language

- Invented in AT&T labs - named after C
- S-Plus vs R (aka 'GNU S')
- Language first and foremost
```{r}
print ("Hello world")
```

## Everything that exists is an object. Everything that happens is a function call
### John Chambers

## We will focus on 

1. **Objects** - *vectors*, matrices, arrays, lists, *dataframes*
2. **Functions** - mean(), sum(), glm()

## Some R code

```{r}
# create two objects, vectors
age <- c(34,55,60,75)
retire <- c(33L,5L,1L,10L)
# do some things with them
(age.retire <- age + retire)
mean(age)
mean(age.retire)
```

## Objects

**Everything** in R is an object.
age and age.retire are atomic vectors, a type of object
```{r}
class (age)
class (retire)
```
But so is mean, even though it is also a function
```{r}
mean2 <- mean
typeof(mean2)
```

## Functions

**Everything** you do in R you do with a function
```{r}
mean(1:3)
is.function (mean)
```
but even + is really a function
```{r}
`+`(2,3)
is.function (`+`)
```

## Our focus

### Understanding objects

### Understanding how functions interact with objects