---
title: 'Lecture 5: Subsetting other objects'
author: "David McAllister"
date: "Monday 31st October, 2016"
output:
  ioslides_presentation: default
---

## Objective of section
* To use subsetting on lists, matrices and dataframes
* To describe how R changes object type with subsetting

## Much of the following is taken from
Advanced R by Hadley Wickham
"http://adv-r.had.co.nz/"

# Lists

## Subsetting lists
Similar, but lists contain objects

```{r}
bigworm <- list (segment1 = 1:20, segment2 = letters[1:10])
str(bigworm)
str(bigworm[1]) # single hooks chop the worm, now got another worm
str(bigworm[[1]]) # second hook to slice it open
str(bigworm$segment1) # dollar sign is double hooks wrought together
```

## Can unlist, but ...
```{r}
unlist (bigworm)
```

## Lists are recursive, metaphor breaks down
```{r, eval = -4}
outside.list <- list (inside.list1 = list (vector1 = 1:20,
                                           vector2 = letters[1:10]),
                      inside.list2 = list (7,8,9))
# str(inside.list1) # R can't see it
str(outside.list)
```

## Absurd structures, eg a list with a single list inside
```{r}
str(outside.list[1])
str(outside.list[[1]])
```

## Complicated nesting
```{r}
str(outside.list$inside.list1$vector1)
str(outside.list$inside.list2)

# Synonymously
str(outside.list[[1]][[1]])
```

## What's the point?

Functions produce single objects

* complex functions produce lists
* Need to use lists
    * Create functions
    * extract from functions
* Dataframes are lists
    * usually composed entirely of vectors, but occassionally not

# Matrices

## Matrices
Simple generalisation of vector subsetting
```{r}
a <- matrix (1:25, nrow = 5, dimnames = list (paste ("rows",1:5), 
                                              paste ("cols",1:5)))

# Integer subsetting
a [1,1]
a [1,]
a [,1]
```

## Matrices continued
```{r}
# Logical subsetting
a [ c(T,F,F,F,F) ,c(T,T,F,F,F)]

# Character subsetting
a [ c("rows 1", "rows 2"), c("cols 1", "cols 2")]
a [ c("rows 1", "rows 1"), ]

```

# Dataframes

## Dataframes, mixture of both

```{r}
head(iris)
dim (iris)
# Integer
iris [2, 1]
iris [2,]

# Character
iris [1:3, "Sepal.Width"]

# Logical - very powerful
iris [iris$Species == "setosa" & iris$Sepal.Width >= 4, c("Sepal.Length", "Sepal.Width")]

```

## Dataframes and dropping-down
```{r}
str(iris$Sepal.Length)
str(iris[,1])
str(iris [[1]])
str(iris [1])
```

## Summary of operators

|             | Simplifying               | Preserving                                   | Simplification
|-------------|---------------------------|----------------------------------------------| ------------
| Vector      | `x[[1]]`                  | `x[1]`                                       | Loses names
| List        | `x[[1]]`                  | `x[1]`                                       | *Shells* list
| Factor      | `x[1:4, drop = T]`        | `x[1:4]`                                     | Loses labels
| Array       | `x[1, ]` __or__ `x[, 1]`  | `x[1, , drop = F]` __or__ `x[, 1, drop = F]` | Becomes vector
| Data frame  | `x[, 1]` __or__ `x[[1]]`  | `x[, 1, drop = F]` __or__ `x[1]`             | Becomes vector

Modified from Hadley Wickham