---
title: 'Lecture 2: Objects'
author: "David McAllister"
date: "Monday 31st October, 2016"
output:
  ioslides_presentation: default
---

## Objective of session

* Describe the different types of R objects
* Describe how R coerces different types of atomic vectors


## Much of the following is taken from
Advanced R by Hadley Wickham
http://adv-r.had.co.nz/

## Types of data object

|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

  
## Atomic vectors

* Logical, integer, double (numeric) and character

* Logical > integer > double > character

* Must be the same type

* created with c() for combine

## Quiz

What will you get if you do the following ...
```{r, eval=FALSE}
c(1L, 6L, 10L)

c(1.1, 2.1, 4.5)

c(1, 2, 4.1, "missing")

c(TRUE, FALSE, 1)
```
When try to combine different types changes vector type
**runs down the hill**
  
## Lists versus vectors
``` {r}
object1 <- c (1, 2, 3)
object2 <- c ("a", "b", "c", "d", "e")
c(object1, object2)
list1 <- list (object1, object2)
str (list1)
```

## Lists are recursive
``` {r}
list2 <- list (list1, list1)
str(list2)
list3 <- c (list1, list1)
str(list3)
```

## Matrices
2d extension of vectors
```{r}
mymatrix <- matrix (1:25, nrow = 5)
print (mymatrix)
dim (mymatrix)
```

## Data frames
* Type of list
    + Allows multiple types; eg character, double, logical
    + Has dimensions
```{r}
head (iris)
dim(iris)
```

## Create data frame
```{r}
mydata <- data.frame (people = c("Sarah", "Nynke", "Colin", "Stephanie"),
                      gender = c(0,0,1,0),
                      stringsAsFactors = FALSE)
mydata
str( as.list (mydata))
```

## Names

```{r}
# Vectors can have names
c(one = 1, two = 2)
# Matrices have colnames and rownames
matrix (1:4, nrow = 2, dimnames = list(c("one", "two"), c("A", "B")))
# Dataframes have names
names(iris)
```

## Same function, different action

```{r, echo = -3}
a_vector <- rep(1:5, 2)
a_matrix <- matrix (a_vector, nrow = 5)
par(mfcol = c(1,2))
plot(a_vector)
plot(a_matrix)
```
