---
title: "Background to R"
author: "David McAllister"
date: "Monday 31st October, 2016"
output:
  ioslides_presentation: default
---

## Aims and objectives
The aim of this course is to have a better feel for how R works as a language. At the end of this course you should be able to

* Clean and transform data in R
* describe how R's objects work
* Understand from the help menu how functions in base R and the packages work
* Use the apply family of functions
* Write simple functions

## Overview
Lectures followed by practicals

### Day 1
R Objects, subsetting, assigning

### Day 2
Attributes, functions, apply family of functions

## The S programming language

- Invented in AT&T labs - named after C
- S-Plus vs R (aka 'GNU S')
- Language first and foremost
```{r}
print ("Hello world")
```

## Everything that exists is an object. Everything that happens is a function call
### John Chambers

## We will focus on 

1. **Objects** - *vectors*, matrices, arrays, lists, *dataframes*
2. **Functions** - mean(), sum(), glm()

## Some R code

```{r}
# create two objects, vectors
age <- c(34,55,60,75)
retire <- c(33L,5L,1L,10L)
# do some things with them
(age.retire <- age + retire)
mean(age)
mean(age.retire)
```

## Objects

**Everything** in R is an object.
age and age.retire are atomic vectors, a type of object
```{r}
class (age)
class (retire)
```
But so is mean, even though it is also a function
```{r}
mean2 <- mean
typeof(mean2)
```

## Functions

**Everything** you do in R you do with a function
```{r}
mean(1:3)
is.function (mean)
```
but even + is really a function
```{r}
`+`(2,3)
is.function (`+`)
```

## Our focus

### Understanding objects

### Understanding how functions interact with objects